The effects of alignment error and alignment filtering on the \sw detection of positive selection

### 2011-09-12 projects/slrsim/paper/slrsim.tex

\documentclass{article}
\usepackage{natbib,graphicx,dblfloatfix,multirow,array,threeparttable}
\usepackage{xspace,bm,amsmath,amssymb}
\usepackage[nolists,tablesfirst,nomarkers]{endfloat}
\usepackage[a4paper,hmargin=2.5cm,vmargin=2.5cm]{geometry}
\usepackage{color}
%\usepackage{ulem}

@@

\newcommand{\mtwo}{PAML M2a\xspace}
\newcommand{\meight}{PAML M8\xspace}
\newcommand{\pranka}{PRANK$_{\textrm{AA}}$\xspace}
\newcommand{\prankc}{PRANK$_{\textrm{C}}$\xspace}
\newcommand{\omg}{\bm{\omega}\xspace}
\newcommand{\tpr}{TPR$_{1\%}$\xspace}
\newcommand{\sw}{sitewise\xspace}
\newcommand{\Sw}{Sitewise\xspace}
\newcommand{\Dr}{{\emph{Drosophila}}\xspace}

\title{The effects of alignment error and alignment filtering on the \sw detection of positive selection}
%%\title{The effects of alignment error and alignment filtering on the \sw detection of positive selection in mammals}
%\title{The effects of alignment error and alignment filtering on the \sw detection of positive selection in vertebrate genomics}
\author{{Gregory Jordan and Nick Goldman}}
%\affiliation{European Bioinformatics Institute, Wellcome Trust Genome Campus, Hinxton, Cambridgeshire, CB10 1SD, United Kingdom}}
\maketitle
%\authorlist{Jordan and Goldman}
%\shorttitle{Effect of Alignment Error on Detecting Sitewise Selection}

%\mbeabstract{% 
\section*{Abstract}
\textcolor{red}{
Alignment filters are commonly applied to try to reduce errors
resulting from misalignment, but the prevalence of such errors and the
ability of filters to reduce them is not well understood.
% 
Focusing on the sitewise detection of positive selection across a wide
range of divergence levels and indel rates, we performed simulation
experiments to quantify the false positives and false negatives
introduced by alignment error and the ability of alignment filters to
improve performance.
%
We found that some aligners led to many false positives while others
resulted in very few.  False negatives were a problem for all
aligners, increasing with divergence. Of the aligners tested, PRANK's
codon-based alignments consistently performed the best and ClustalW
performed the worst.
%
Of the filters tested, GUIDANCE performed the best and Gblocks
performed the worst. Although some filters showed good ability to
reduce the error rates from ClustalW and MAFFT alignments, none were
found to substantially improve the performance of PRANK alignments
under most \textcolor{red}{conditions}.
% 
Our results revealed distinct trends in error rates and power levels
for aligners and filters within a biologically plausible parameter
space. Controls using the true alignment and an optimal filtering
method suggested that performance improvements could be gained by
improving aligners or filters to reduce the prevalence of false
negatives, especially at higher divergence levels and indel rates.
}
%\keywords{Alignment, positive selection, alignment filtering, indels}

@@

Reflecting a widespread awareness of the problem of misalignment,
methods for identifying and removing uncertain or unreliable alignment
regions have been commonly used in phylogenetic and molecular
evolutionary analyses. The popular Gblocks program applies a set of
heuristic criteria to identify conserved blocks deemed suitable for
phylogenetic or evolutionary analysis \citep{Castresana2000Selection}
while a number of aligners such as T-Coffee
\citep{Notredame2000TCoffee} and PRANK
\citep{Loytynoja2008PhylogenyAware} produce estimates of alignment
confidence or reliability. GUIDANCE, which measures the robustness of
alignment regions to perturbations in the guide tree used for
progressive alignment, has also been proposed as an alignment
confidence score \citep{Penn2010Alignment}. Unfortunately, despite
their widespread use, the impact of the many available alignment
scoring and filtering methods on phylogenetic and evolutionary
analyses has not been well studied. Even for a single filtering
program, Gblocks, results have been contradictory: one
simulation-based study found that it improved the phylogenetic signal
\citep{Talavera2007Improvement} while an empirical study across a wide
range of taxa found that Gblocks-filtered alignments produced worse
phylogenetic trees than unfiltered alignments
\citep{Dessimoz2010Phylogenetic}. With the application of published
filtering methods to alignments before testing for positive selection
becoming standard practice
\citep{Studer2008Pervasive,Aguileta2009Rapidly}, an analysis of the
benefits of alignment filtering to the detection of positive selection
would seem well-warranted.

This paper aims to use a simulation framework to incorporate alignment
error and alignment filtering into estimates of the error rate and
power of \sw evolutionary inference of positive selection. Our
approach builds from that of Anisimova et al.
\citeyearpar{Anisimova2002Accuracy} in that we use simulated protein
alignments to evaluate methods for detecting \sw positive selection,
considering each site to be an independent hypothesis test producing a
positive or negative result. \textcolor{red}{As we simulate codon
  alignments with indels and calculate inferred alignments using a
  small but diverse sample of aligners and alignment filters, the
  experimental design also shares some features with Fletcher and
  Yang's \citeyearpar{Fletcher2010Effect} investigation of the
  performance of the branch-site method for detecting positive
  selection under alignment error and Privman et al.'s
  \citeyearpar{Privman2011Improving} study of the effect of alignment
  filtering on detecting positive selection in HIV-1 genes. However,
  the present work differs from previous studies in a number of
  important ways.}

\textcolor{red}{Unlike Fletcher and Yang, here we focus on \sw
  detection of sitewise positive selection occurring throughout a
  phylogeny (as opposed to genewise detection of selection acting at
  specific branches) and we evaluate the impact of a number of
  alignment filtering methods on the \sw analysis.}

@@

\textcolor{red}{We hypothesize that these factors may strongly affect
  the the performance of methods for alignment and detection of
  selection. Accordingly, our simulations employ parameters similar to
  those commonly encountered in the \sw analysis of vertebrate gene
  families and encompass a wide range of biologically plausible indel
  rates and divergence levels. Although our main purpose is to
  understand the behaviour of various aligners and filters within our
  range of simulations, we will make use of the above-noted
  methodological similarities by pointing out those aspects of our
  study which agree with or contradict the literature where possible.}

@@

The distribution of \sw selective pressures (embodied by the parameter
$\omg$, the ratio of the rate of non-synonymous substitution to the
rate of synonymous substitution) was modeled with a log-normal
distribution derived from a maximum-likelihood fit to a large dataset
of \sw selective pressures estimated from mammalian gene trees
(\citeyearpar{LindbladToh2011High}; log-normal parameters shown in Table
\ref{table_1}). This distribution, with mean $\omg$ of 0.28 and 6\% of
sites having $\omg>1$, is consistent with the structure-based
expectation of many protein sites under purifying selection and few
under neutral selection or positive selection
\citep{Smith1970Natural,Kimura1974SomePrinciples}. INDELible's general
discrete model of \sw $\omg$ variation was used to approximate the
log-normal distribution by splitting the probability density into 50
equally-spaced bins between $\omg$ values of 0 and 3, with the highest
bin containing the probability density for all values $\omg>3$.

@@

When alignments were inferred using one of the three aligners tested,
the TPR, \tpr{} and $\omg$ correlation were all reduced relative to
the true alignment (dashed and dotted lines, Figure \ref{fig_3}). The
degree of reduction varied depending on the aligner, simulation
conditions, and performance measurement being analyzed. At low
divergences (e.g. MPL $<$ 0.2) the inferred alignments generally
showed only a small decrease in performance. As divergence levels
increased, so did the difference between the performance of the true
alignment and the inferred alignments. The three aligners tested could
be consistently and unambiguously ranked by all of the measured
performance characteristics, with \prankc always performing best and
ClustalW performing worst. The same ranking of aligners with respect
to detecting positive selection was found by Fletcher and Yang
\citeyearpar{Fletcher2010Effect}; our results corroborate their
finding and provide evidence that this ranking may be applicable to a
wider array of evolutionary analyses.

@@

As alignment filters act through the removal of alignment residues or
columns, a certain amount of reduction in the FPR and TPR was expected
purely from the decreased amount of information available. For
example, a filter that randomly removes a fraction of residues of each
alignment would be expected to produce equal reductions in FPR and
TPR. A more effective filter may also yield a reduced TPR, but the FPR
reduction would be larger in magnitude, making the detection of
positive selection more powerful for a given error rate. Thus, a
reduced FPR is not necessarily indicative of good filtering
performance, nor is a reduced TPR necessarily indicative of poor
filtering performance. Additionally, the prevalence of false negatives
resulting from misalignment suggested that alignment filters may also
improve power by removing false negatives, perhaps by masking out
residues that were preventing positive sites from being
identified. The removal of false negatives would result in an
increased TPR, further complicating the assessment of filtering
results based on FPR or TPR alone. As a result, we focused on the
change in error-controlled \tpr{} as the best single measure of
whether a filter had successfully improved the \sw power of a dataset
since this value is sensitive to changes in both the FPR and TPR. Note
that the \tpr controls the FPR post-filtering, accounting for the
tendency of filtering to reduce the FPR at a given cutoff threshold.

We first examined the two controls, the unfiltered true alignment and
the inferred alignments filtered with the optimal filter (top two
rows, Figure \ref{fig_5}). The true alignment nearly always showed
smaller FPR (red cells) and greater TPR and \tpr (blue cells) compared
to the inferred alignments, with a greater magnitude of change
relative to the ClustalW alignments than to the \prankc alignments
(darker shades cf. lighter shades). These scores represented the
direction and an upper limit on the magnitude of change that might be
achieved by a perfect alignment filter. One exception to the general
trend was the observation of two simulation conditions with slightly
elevated FPRs in the true alignment compared to \prankc alignments (at
an indel rate of 0.04 and MPL of 0.8 and 2.0). This small
inconsistency may be explained by stochastic variation in false
positive counts, as the absolute value of the FPR was very low in both
datasets under those conditions (on the order of $5\times10^{-4}$,
Figure \ref{fig_4}B). A similar slight FPR elevation was also observed
at the same indel rate for the optimal, GUIDANCE and T-Coffee filters.

@@

Filtering was less beneficial when applied to the more accurate
\prankc alignments, with T-Coffee reducing performance and GUIDANCE
yielding only mild \tpr improvements. Importantly, GUIDANCE only
showed improved performance at high divergence levels (e.g.,
MPL$>$1.6), well above those found in commonly-analyzed groups of
species. Thus, the use of unfiltered \prankc alignments would yield
the best performance for detecting \sw positive selection when
analyzing protein-coding sequences across a wide range of indel rates
and commonly encountered sequence divergence levels. The performance
of the `optimal' filter on \prankc alignments suggests that mild
further improvements to filtering strategies may be possible, but the
potential for improvement is small and may be of little value.

@@

In this paper, we investigated the performance of sitewise detection
of positive selection under a range of tree sizes, indel rates, and
divergence levels, using simulation parameters designed to approximate
the analysis of mammalian and other typical protein-coding genes. We
evaluated the ability of four alignment methods and three alignment
filtering methods to produce alignments for detecting positively
selected sites, using the FPR, TPR, and error-controlled \tpr of the
sitewise detection of positive selection as our main measures of
performance.

The simulation results showed that alignment error can have a
measurable impact on the error rates and power of the sitewise
detection of positive selection under all but the least difficult
alignment conditions. We confirmed and extended the findings of
Fletcher and Yang \citeyearpar{Fletcher2010Effect} regarding the
relative accuracy of different aligners, showing that \prankc had the
best performance and ClustalW had the worst performance for subsequent
sitewise analysis. Notably, our simulations found that ClustalW
produced more sitewise false positives than MAFFT or \prankc even at
low MPLs, suggesting that its use should be avoided even when
analyzing closely-related sequences. \prankc, on the other hand,
resulted in very low FPRs even at higher MPLs. In particular, when the
number of sequences in the tree was large, \prankc{}'s sitewise FPRs
were virtually indistinguishable from those of the true alignment.

@@

The manual or automated adjustment of alignments has been thought by
many to be an important step in evolutionary analyses due to fear of a
high prevalence of misalignment-induced false positives. While this is
true for some aligners, we find that more accurate alignment
algorithms result in significantly fewer false positives in the
subsequent detection of sitewise positive selection. This strongly
reduces the beneficial effect of alignment filtering, so much so that
current filtering methods are essentially unable to improve the
performance of \prankc alignments when analyzed with SLR.

As a result, among the aligners and filters tested here, we recommend
the use of unfiltered \prankc alignments for the detection of sitewise
positive selection in all but the most divergent phylogenies. GUIDANCE
showed some ability to improve sitewise power under difficult
alignment conditions at high divergence levels, but this improvement
was achieved largely through the recovery of false negatives as
opposed to the elimination of false positives. For an analysis where
the control of false positives is the primary concern, the added
computational expense of running many bootstrap alignment replicates
(as performed by GUIDANCE) may not be worth the possibility of a
slight increase in power. The development of efficient and effective
alignment filters, particularly with an eye towards recovering false
negatives as identified in this study, will be an interesting area for
future research.

Our simulations did not include fully biologically realistic models of
spatial or temporal variability in the rate of indel formation or in
the distribution of selective pressures \citep{Whelan2008Spatial}, but
we do not expect that such heterogeneity would affect our main
conclusions regarding the relative performance of different aligners
and filters. The trends we observed were consistent across a wide
range of parameter values, suggesting that they reflect fundamental
differences in each method's ability to align or filter sequences as
opposed to artifacts due to the relative simplicity of our
simulations.

@@

GJ was funded by a Gates Cambridge Trust Scholarship and is a member
of Darwin College, University of Cambridge.

>>>>

\documentclass{article}
\usepackage{natbib,graphicx,dblfloatfix,multirow,array,threeparttable}
\usepackage{xspace,bm,amsmath,amssymb}
\usepackage[nolists,tablesfirst,nomarkers]{endfloat}
\usepackage[a4paper,hmargin=2.5cm,vmargin=2.5cm]{geometry}
\usepackage{color}

@@

\newcommand{\mtwo}{PAML M2a\xspace}
\newcommand{\meight}{PAML M8\xspace}
\newcommand{\pranka}{PRANK$_{\textrm{AA}}$\xspace}
\newcommand{\prankc}{PRANK$_{\textrm{C}}$\xspace}
\newcommand{\omg}{\bm{\omega}\xspace}
\newcommand{\tpr}{TPR$_{1\%}$\xspace}
\newcommand{\tprf}{TPR$_{5\%}$\xspace}
\newcommand{\sw}{sitewise\xspace}
\newcommand{\Sw}{Sitewise\xspace}
\newcommand{\Dr}{{\emph{Drosophila}}\xspace}

\title{The effects of alignment error and alignment filtering on the \sw detection of positive selection}
%%\title{The effects of alignment error and alignment filtering on the \sw detection of positive selection in vertebrates}
\author{{Gregory Jordan and Nick Goldman}}
%\affiliation{European Bioinformatics Institute, Wellcome Trust Genome Campus, Hinxton, Cambridgeshire, CB10 1SD, United Kingdom}}
\maketitle
%\authorlist{Jordan and Goldman}
%\shorttitle{Effect of Alignment Error on Detecting Sitewise Selection}

%\mbeabstract{% 
\section*{Abstract}
\textcolor{red}{
Alignment filters are commonly applied to try to reduce errors
resulting from misalignment, but the prevalence of such errors and the
ability of filters to reduce them is not well understood.
% 
Focusing on the sitewise detection of positive selection across a wide
range of divergence levels and indel rates, we performed simulation
experiments to quantify the false positives and false negatives
introduced by alignment error and the ability of alignment filters to
improve performance.
%
We found that some aligners led to many false positives while others
resulted in very few.  False negatives were a problem for all
aligners, increasing with divergence. Of the aligners tested, PRANK's
codon-based alignments consistently performed the best and ClustalW
performed the worst.
%
Of the filters tested, GUIDANCE performed the best and Gblocks
performed the worst. Although some filters showed good ability to
reduce the error rates from ClustalW and MAFFT alignments, none were
found to substantially improve the performance of PRANK alignments
under most \textcolor{red}{conditions}.
% 
Our results revealed distinct trends in error rates and power levels
for aligners and filters within a biologically plausible parameter
space, and controls using the true alignment and an optimal filtering
method suggested that performance improvements could be gained by
improving aligners or filters to reduce the prevalence of false
negatives, especially at higher divergence levels and indel rates.
}
%\keywords{Alignment, positive selection, alignment filtering, indels}

@@

Reflecting a widespread awareness of the problem of misalignment,
methods for identifying and removing uncertain or unreliable alignment
regions have been commonly used in phylogenetic and molecular
evolutionary analyses. The popular Gblocks program applies a set of
heuristic criteria to identify conserved blocks deemed suitable for
phylogenetic or evolutionary analysis \citep{Castresana2000Selection}
while a number of aligners such as T-Coffee
\citep{Notredame2000TCoffee} and PRANK
\citep{Loytynoja2008PhylogenyAware} produce estimates of alignment
confidence or reliability. GUIDANCE, which measures the robustness of
alignment regions to perturbations in the guide tree used for
progressive alignment, has also been proposed as an alignment
confidence score \citep{Penn2010Alignment}. Unfortunately, despite
their widespread use, the impact of the many available alignment
scoring and filtering methods on phylogenetic and evolutionary
analyses has not been well studied. Even for a single filtering
program, Gblocks, results have been contradictory: one
simulation-based study found that it improved the phylogenetic signal
\citep{Talavera2007Improvement} while an empirical study across a wide
range of taxa found that Gblocks-filtered alignments produced worse
phylogenetic trees than unfiltered alignments
\citep{Dessimoz2010Phylogenetic}. \textcolor{red}{A recent study suggested that the
benefit of alignment filtering (in terms of improved accuracy)
outweighs the cost (in terms of reduced power) when applied to
detecting positive selection, but the analysis was limited to a small
range of possible evolutionary scenarios, as discussed below
\citep{Privman2011Improving}.}  With the application of published
filtering methods to alignments before testing for positive selection
becoming standard practice
\citep{Studer2008Pervasive,Aguileta2009Rapidly}, \textcolor{red}{continued investigation into the}
potential benefits of alignment filtering to the detection of positive selection
would seem well-warranted.

This paper aims to use a simulation framework to incorporate alignment
error and alignment filtering into estimates of the error rate and
power of \sw evolutionary inference of positive selection. Our
approach \textcolor{red}{is modeled on that of} Anisimova et al.
\citeyearpar{Anisimova2002Accuracy} in that we use simulated protein
alignments to evaluate methods for detecting \sw positive selection,
considering each site to be an independent hypothesis test producing a
positive or negative result. \textcolor{red}{As we simulate codon
  alignments with indels and calculate inferred alignments using a
  small but diverse sample of aligners and alignment filters, the
  experimental design also shares some features with Fletcher and
  Yang's \citeyearpar{Fletcher2010Effect} investigation of the
  performance of the branch-site method for detecting positive
  selection under alignment error and Privman et al.'s
  study of the effect of alignment
  filtering on detecting positive selection in HIV-1 genes. However,
  the present work differs from previous studies in a number of
  important ways.}

\textcolor{red}{Unlike Fletcher and Yang, here we focus on \sw
  detection of sitewise positive selection occurring throughout a
  phylogeny (as opposed to genewise detection of selection acting at
  specific branches) and we evaluate the impact of a number of
  alignment filtering methods on the \sw analysis. Thus, the
  biological hypothesis being investigated is different, and the
  addition of filtering methods adds another layer of complexity to
  our analysis.}

@@

\textcolor{red}{We hypothesize that the divergence level and indel
  rate, two highly variable and important evolutionary factors, may
  strongly affect the performance of methods for alignment and
  detection of selection. Accordingly, our simulations encompass a
  wide range of biologically plausible indel rates and divergence
  levels while fixing other parameters at values typical to those
  encountered in the \sw analysis of vertebrate gene
  families. Although our main purpose is to understand the behaviour
  of various aligners and filters within our range of simulations, we
  will make use of the above-noted methodological similarities in
  order to highlight aspects of our results which agree with or
  contradict the literature where possible.}

@@

The distribution of \sw selective pressures (embodied by the parameter
$\omg$, the ratio of the rate of non-synonymous substitution to the
rate of synonymous substitution) was modeled with a log-normal
distribution derived from a maximum-likelihood fit to a large dataset
of \sw selective pressures estimated from mammalian gene trees
(Lindblad-Toh et al. \citeyear{LindbladToh2011High}; log-normal parameters shown in Table
\ref{table_1}). This distribution, with mean $\omg$ of 0.28 and 6\% of
sites having $\omg>1$, is consistent with the structure-based
expectation of many protein sites under purifying selection and few
under neutral selection or positive selection
\citep{Smith1970Natural,Kimura1974SomePrinciples}. INDELible's general
discrete model of \sw $\omg$ variation was used to approximate the
log-normal distribution by splitting the probability density into 50
equally-spaced bins between $\omg$ values of 0 and 3, with the highest
bin containing the probability density for all values $\omg>3$.

@@

When alignments were inferred using one of the three aligners tested,
the TPR, \tpr{} and $\omg$ correlation were all reduced relative to
the true alignment (dashed and dotted lines, Figure \ref{fig_3}). The
degree of reduction varied depending on the aligner, simulation
conditions, and performance measurement being analyzed. At low
divergences (e.g. MPL $<$ 0.2) the inferred alignments generally
showed only a small decrease in performance. As divergence levels
increased, so did the difference between the performance of the true
alignment and the inferred alignments. The three aligners tested could
be consistently and unambiguously ranked by all of the measured
performance characteristics, with \prankc always performing best and
ClustalW performing worst. \textcolor{red}{The same ranking of
  aligners with respect to detecting positive selection has been
  observed in a number of studies
  \citep{Fletcher2010Effect,Markova-Raina2011High,Privman2011Improving};
  our results corroborate these findings and provide evidence that
  this ranking may be applicable across a wide range of divergence
  levels and indel rates.}

@@

As alignment filters act through the removal of alignment residues or
columns, a certain amount of reduction in both the FPR and TPR was expected
purely from the decreased amount of information available. For
example, a filter that randomly removes a fraction of residues of each
alignment would be expected to produce equal reductions in FPR and
TPR. A more effective filter may also yield a reduced TPR, but the FPR
reduction would be larger in magnitude, making the detection of
positive selection more powerful for a given error rate. Thus, a
reduced FPR is not necessarily indicative of good filtering
performance, nor is a reduced TPR necessarily indicative of poor
filtering performance. Additionally, the prevalence of false negatives
resulting from misalignment suggested the interesting possibility that alignment filters may also
improve power by removing false negatives, perhaps by masking out
residues that were preventing positive sites from being
identified. The removal of false negatives would result in an
increased TPR, further complicating the assessment of filtering
results based on FPR or TPR alone. As a result, we focused on the
change in error-controlled \tpr{} as the best single measure of
whether a filter had successfully improved the \sw power of a dataset
since this value is sensitive to changes in both the FPR and TPR. Note
that the \tpr controls the FPR post-filtering, accounting for the
tendency of filtering to reduce the FPR at a given cutoff threshold.

We first examined the two controls, the unfiltered true alignment and
the inferred alignments filtered with the optimal filter (top two
rows, Figure \ref{fig_5}). The true alignment nearly always showed
smaller FPR (red cells) and greater TPR and \tpr (blue cells) compared
to the inferred alignments, with a greater magnitude of change
relative to the ClustalW alignments than to the \prankc alignments
(darker shades cf. lighter shades). These scores represented the
direction and an upper limit on the magnitude of change that might be
achieved by a perfect alignment filter. One exception to the general
trend of lower FPR in the true alignment was the observation of two
simulation conditions with slightly elevated FPRs in the true
alignment compared to \prankc alignments (at an indel rate of 0.04 and
MPL of 0.8 and 2.0). This small inconsistency may be explained by
stochastic variation in false positive counts, as the absolute value
of the FPR was very low in both datasets under those
conditions. \textcolor{red}{Figure \ref{fig_4}B shows the FPR to be on
  the order of $5\times10^{-4}$ under those conditions, and we
  observed 63 false positives for the true alignment and 67 false
  positives for the \prankc alignment across all 150 replicates at the
  indel rate of 0.04 and MPL of 0.8}. A similar slight FPR elevation
was also observed at the same indel rate for the optimal, GUIDANCE and
T-Coffee filters.

@@

Filtering was less beneficial when applied to the more accurate
\prankc alignments, with T-Coffee reducing performance and GUIDANCE
yielding only mild \tpr improvements. Importantly, GUIDANCE only
showed improved performance at high divergence levels (e.g.,
MPL$>$1.6), well above those found in commonly-analyzed groups of
species. \textcolor{red}{Thus, the use of unfiltered \prankc alignments would yield
largely equivalent performance to GUIDANCE-filtered alignments for detecting
\sw positive selection when analyzing protein-coding sequences across
a wide range of indel rates and commonly encountered sequence
divergence levels. On the other hand, GUIDANCE (when judiciously applied, including an upper limit on the amount of sequence data removed) never significantly reduced performance compared to the unfiltered alignments.} Finally, we note that the performance of the `optimal' filter on \prankc
alignments suggests that mild further improvements to filtering
strategies may be possible, but the potential for improvement is small
and may be of little value.

@@

In this paper, we investigated the performance of sitewise detection
of positive selection under a range of tree sizes, indel rates, and
divergence levels, using simulation parameters designed to approximate
the analysis of typical mammalian protein-coding genes. We
evaluated the ability of four alignment methods and three alignment
filtering methods to produce alignments for detecting positively
selected sites, using the FPR, TPR, and error-controlled \tpr of the
sitewise detection of positive selection as our main measures of
performance.

The simulation results showed that alignment error can have a
measurable impact on the error rates and power of the sitewise
detection of positive selection under all but the least difficult
alignment conditions. We confirmed and extended the findings of
Fletcher and Yang \citeyearpar{Fletcher2010Effect}, Markova-Raina et al. \citeyearpar{Markova-Raina2011High}, and Privman et al. \citeyearpar{Privman2011Improving} regarding the
relative accuracy of different aligners, showing that \prankc had the
best performance and ClustalW had the worst performance for subsequent
sitewise analysis. Notably, our simulations found that ClustalW
produced more sitewise false positives than MAFFT or \prankc even at
low MPLs, suggesting that its use should be avoided even when
analyzing closely-related sequences. \prankc, on the other hand,
resulted in very low FPRs even at higher MPLs. In particular, when the
number of sequences in the tree was large, \prankc{}'s sitewise FPRs
were virtually indistinguishable from those of the true alignment.

@@

The manual or automated adjustment of alignments has been thought by
many to be an important step in evolutionary analyses due to fear of a
high prevalence of misalignment-induced false positives. While this is
true for some aligners, we find that more accurate alignment
algorithms result in significantly fewer false positives in the
subsequent detection of sitewise positive selection. This strongly
reduces the beneficial effect of alignment filtering, so much so that
current filtering methods are \textcolor{red}{scarcely able to} improve the
performance of \prankc alignments when analyzed with SLR.

As a result, we cannot unequivocally recommend the use of alignment
filtering in the detection of sitewise positive selection, except
perhaps for the most divergent and indel-prone sequences, which are
unlikely to be encountered in analyses focusing on mammalian or
vertebrate genes. Although GUIDANCE showed some ability to improve the
error-controlled power under difficult alignment conditions at high
divergence levels (whereas T-Coffee and Gblocks failed to improve upon
any \prankc alignments) this improvement was modest in magnitude and
was achieved largely through the recovery of false negatives as
opposed to the elimination of false positives. For an analysis where
the control of false positives is the primary concern, the added
computational expense of running many bootstrap alignment replicates
(as performed by GUIDANCE) may not be worth the possibility of a
slight increase in power. On the other hand, judicious application of
GUIDANCE never significantly reduced power, so its use would not be
expected to yield worse results.

\textcolor{red}{Some of our results and conclusions differed from those of Privman et
al. \citeyearpar{Privman2011Improving}, who found strong improvements
in error-controlled power by filtering alignments in simulations
focused on three HIV-1 genes. Although the phylogenetic trees they
used to guide their simulations contained divergence levels at the low
end of our tested range (MPLs of 0.38, 0.34 and 0.33 for gag, pol and
env, respectively; E. Privman, personal communication) and were roughly
comparable in size and shape to our 17-taxon tree, we failed to find a
significant benefit for alignment filters at any MPL below 1.2 when
using \prankc alignments. Interestingly, the authors also found
Gblocks to be roughly comparable in performance to GUIDANCE, while we
found Gblocks' performance to be very poor. Some of these
discrepancies may be due to differences in the details of our
simulations or filtering procedures, but in the end our results are
largely complementary: Privman et al. showed that filtering can be
beneficial for detecting positive selection, especially in the case of
fast-evolving (but fairly closely-related) sequences, while we have
shown that divergence levels and indel rates have a significant impact
on the accuracy and sensitivity of different aligners.}

Our simulations did not include fully biologically realistic models of
spatial or temporal variability in the rate of indel formation or in
the distribution of selective pressures \citep{Whelan2008Spatial}, but
we do not expect that such heterogeneity would affect our main
conclusions regarding the relative performance of different aligners
and filters. The trends we observed were consistent across a wide
range of parameter values and tree sizes, suggesting that they reflect
fundamental differences in each method's ability to align or filter
sequences as opposed to artifacts due to the relative simplicity of
our simulations.

@@

The authors would like to thank Eyal Privman for providing HIV-1 tree
data and useful discussion. GJ was funded by a Gates Cambridge Trust
Scholarship and is a member of Darwin College, University of
Cambridge.

### 2011-09-12 projects/slrsim/paper/supp_slrsim.tex

\newcommand{\pranka}{PRANK$_{\textrm{AA}}$\xspace}

@@

\begin{figure}[t]
\begin{center}
%\includegraphics[scale=0.65]{supp_fig1.pdf}
\includegraphics[scale=0.8]{supp_fig1.pdf}
\end{center}
\caption{This figure depicts the same simulations and uses the same
  formatting as Figure 4, except that results for MAFFT and \pranka
  have been added to sections (a) and (b) and results using ClustalW +
  GUIDANCE have been added to all sections.}
\label{supp_fig1}
\end{figure}

>>>>

\newcommand{\mtwo}{PAML M2a\xspace}
\newcommand{\meight}{PAML M8\xspace}
\newcommand{\pranka}{PRANK$_{\textrm{AA}}$\xspace}
\newcommand{\prankc}{PRANK$_{\textrm{C}}$\xspace}
\newcommand{\omg}{\bm{\omega}\xspace}
\newcommand{\tpr}{TPR$_{1\%}$\xspace}
\newcommand{\tprf}{TPR$_{5\%}$\xspace}
\newcommand{\sw}{sitewise\xspace}
\newcommand{\Sw}{Sitewise\xspace}
\newcommand{\Dr}{{\emph{Drosophila}}\xspace}

@@

\begin{figure}[t]
\begin{center}
%\includegraphics[scale=0.65]{supp_fig1.pdf}
\includegraphics[scale=0.7]{supp_fig1.pdf}
\end{center}
\caption{This figure depicts the same simulations and uses the same
  formatting as Figure 4, except that results for MAFFT and \pranka
  have been added to sections (a) and (b) and results using ClustalW +
  GUIDANCE and MAFFT + GUIDANCE have been added to all sections.}
\label{supp_fig1}
\end{figure}

\begin{figure}[t]
\begin{center}
%\includegraphics[scale=0.65]{supp_fig1.pdf}
\includegraphics[scale=0.55]{supp_fig2.pdf}
\end{center}
\caption{This figure depicts largely the same simulations and uses the
  same formatting as Figure 5 except for two changes. First,
  results for MAFFT have been added to section (b) and results for
  \prankc have been moved to section (c). Second, the rightmost column
  has been added, showing the true positive rate (TPR) at a 5\% false
  positive rate (FPR) threshold (labeled \tprf).}
\label{supp_fig2}
\end{figure}
